On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-21 paper/appendix3d.tex
reg udpate

An approximate solution for the growth rate may be obtained for small
$|K|$ by balancing the last two terms,
\begin{align}
  S \simeq \frac{\alpha K^2\left[2|K|/Q- \gamma K^2 + 
     2q(2-q)(\gamma-1)\right]}{2(2-q)  - 2|K|/Q + \gamma K^2 }, 
\end{align}
which coincides with \citeauthor{gammie96}'s Eq. 18 for $\gamma=1$. 
%We recover \cite{gammie96}'s dispersion relation 
%for viscous gravitational instability as follows.  Next we
%simplify Eq. \ref{bigA}---\ref{bigF} by assuming $|\beta S |\gg 1$  
%This will turn out to be equivalent to $\beta\gg 1$, i.e. the
%adiabatic limit. 
%and $S\lesssim O(K^2)$ as $|K|\to 0$. Applying these considerations to
%Eq. \ref{thindisk} gives 
%\begin{align}
%  S = \frac{\alpha K^2\left[\frac{2|K|}{Q} - \gamma K^2 +
%      2q(2-q)(\gamma-1)\right]}{2(2-q) + \gamma K^2 - \frac{2K}{Q}}, 
%\end{align}
%which does not explicitly depend on cooling. 
%Setting $\gamma=1$ then gives \citeauthor{gammie96}'s Eq. 18. The above
%is expression thus generalizes \citeauthor{gammie96}'s result to
%non-isothermal disks.

@@

>>>>

An approximate solution for the growth rate may be obtained for small
$|K|$ by balancing the last two terms,
\begin{align}
  S \simeq \frac{\alpha K^2\left[2|K|/Q- \gamma K^2 + 
     2q(2-q)(\gamma-1)\right]}{2(2-q)  - 2|K|/Q + \gamma K^2 }, 
\end{align}
which coincides with \citeauthor{gammie96}'s Eq. 18 for $\gamma=1$. In
that case the condition for instability is
\begin{align}
%  \frac{2|K|}{Q} - K^2>0. 
  |K|< \frac{2}{Q}.
\end{align} 
This is the same stability condition as if rotation was neglected in
the standard, isothermal dispersion relation for density waves
(obtained from Eq. \ref{thindisk} by setting $\gamma=\theta=1$). Thus
the effect of viscosity is to remove 
rotational stabilization against self-gravity, as originally explained
by \cite{lynden-bell74}.

%We recover \cite{gammie96}'s dispersion relation 
%for viscous gravitational instability as follows.  Next we
%simplify Eq. \ref{bigA}---\ref{bigF} by assuming $|\beta S |\gg 1$  
%This will turn out to be equivalent to $\beta\gg 1$, i.e. the
%adiabatic limit. 
%and $S\lesssim O(K^2)$ as $|K|\to 0$. Applying these considerations to
%Eq. \ref{thindisk} gives 
%\begin{align}
%  S = \frac{\alpha K^2\left[\frac{2|K|}{Q} - \gamma K^2 +
%      2q(2-q)(\gamma-1)\right]}{2(2-q) + \gamma K^2 - \frac{2K}{Q}}, 
%\end{align}
%which does not explicitly depend on cooling. 
%Setting $\gamma=1$ then gives \citeauthor{gammie96}'s Eq. 18. The above
%is expression thus generalizes \citeauthor{gammie96}'s result to
%non-isothermal disks.

@@

\section{Numerical method for the 3D eigenvalue problem}
We use a pseudo-spectral method to solve the set of ordinary
differential equations, Eq. \ref{lin_mass}---\ref{lin_gravity}, on 
the domain $ z\in[0,\zmax]$ with a parity condition at the midplane.
We expand $\bm{U}=[\delta P,\delta\rho,\delta\Phi,
  \delta v_x, \delta 
  v_y]$ in even Chebyshev polynomials,
\begin{align}
  \bm{U}(z) = \sum_{j=1}^N \bm{a}_jT_{2(j-1)}(z/\zmax), 
\end{align}
and the vertical velocity in odd Chebyshev polynomials,
\begin{align}
  \delta v_z(z) = \sum_{j=1}^N b_jT_{2j-1}(z/\zmax),   
\end{align}
where $\bm{a}_j$ and $b_j$ are the spectral coefficients. 
The basis functions are chosen to satisfy a reflecting boundary
condition at the midplane, %GI has even symmetry across midplane 
\begin{align}
  \bm{U}^\prime(0) = \bm{0}, \quad \delta v_z(0) = 0.
\end{align}
%At the upper disk boundary we also apply reflection,
We explicitly apply a reflecting upper disk boundary, 
\begin{align}
  \delta v_z(\zmax) = \delta v_x^\prime(\zmax) = \delta
  v_y^\prime(\zmax) = 0,
\end{align}
and the potential satisfies
\begin{align}
  \delta \Phi^\prime (\zmax) + k \delta\Phi(\zmax) = 0, 
\end{align}
as derived by \cite{goldreich65a} and used in similar studies  
\citep{kim12,lin14c}. %reflection is standard in direct simulations

We discretize the equations, including upper disk boundary conditions,
over the $N$ positive abscissae of the extrema of $T_{2N-1}$. This
procedure converts the differential equations into a generalized
eigenvalue problem, for which we use the standard matrix package
LAPACK to solve. We use $N=65$.

### 2016-01-21 paper/paper3d.tex
reg udpate

\begin{abstract}
 Low mass, self-gravitating accretion disks can remain in a  
quasi-steady, `gravito-turbulent' state provided that thermal 
time-scales are sufficiently long, or equivalently for sufficiently
weak turbulent stresses. Otherwise, numerical simulations show that
disk fragmentation ensues. 
We quantify the interpretation of this result as the
compromise of gravitational stability by: 
\begin{inparaenum}[1)] 
\item 
cooling, which reduces pressure support; and/or
\item 
viscosity, which reduces rotational support. 
\end{inparaenum}
We analyse the gravitational stability of accretion disks
including these non-ideal effects explicitly.   
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. We suggest that realistic disk fragmentation can be
physically attributed,  
at least partially, to the removal of rotational stabilization against
self-gravity by turbulence.   
\end{abstract}
%we suggest turbulent stresses play an active, dynamical role in disk fragmentation by 
%removing rotational stabilization against self-gravity. 
%in the sense of small-scale transport

\section{Introduction}\label{intro} 
The gravitational instability (GI) of rotating disks   
plays a fundamental role in many astrophysical problems. In the context of 
protoplanetary disks (PPDs), GI has two applications. It can 
provide gravitational torques to transport angular momentum outwards
and thus enable mass accretion inwards \citep{armitage10,turner14}.   
GI may also lead to disk fragmentation, which has been invoked to explain planet formation at
large radii \citep{helled14}. Any realistic study of these phenomena require direct
numerical simulations. Nevertheless, physical insight can be  
obtained through analytical understanding of GI.  %physical insight

The standard metric for the (inverse) strength of disk self-gravity is the 
\emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre $Q$ is a
measure of the destabilizing effect of self-gravity ($\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure
($c_s$). When $Q\leq 1$ the disk is formally unstable to local,
axisymmetric disturbances in a two-dimensional (2D, razor-thin),
inviscid and adiabatic (including isothermal) disk.

@@

The Toomre  parameter is widely applied to assess GI in more general 
situtations (cf. the ideal disk models from which $Q$ is derived
from).  This includes realistic disk structures
obtained with non-ideal physics that are not explicitly contained in
$Q$ \citep[e.g.][]{kimura12}.  It is, however, not obvious that such additions
leave the stability condition unaltered. Indeed, we show that GI in more
realistic disk models cannot be described by a single `Toomre-like' 
condition.

%disk structure may be computed with sophisticated physics but
%stability condition still based on ideal disk models. 
%three main ingredients - gravity, pressure, rotation 
%a standard way to apply this condition is to compute the disk
%structure  based on realistic physics

The goal of this work is to generalize the analytic treatment of disk 
GI to allow a self-consistent discussion of GI in realistic accretion
disk models. Specifically, we will examine the effect of cooling and
viscosity on GI in PPDs.  
%Their relevance to PPDs are described below.

>>>>

\begin{abstract}
%Low mass, self-gravitating accretion disks can remain in a  
%quasi-steady, `gravito-turbulent' state provided that thermal 
%time-scales are sufficiently long, or equivalently for sufficiently
%weak turbulent stresses. Otherwise, numerical simulations show that
%disk fragmentation ensues. 
%{\bf
Low mass, self-gravitating accretion disks admit a quasi-steady,
`gravito-turbulent' state in which cooling balances turbulent
heating. However, numerical simulations show that gravito-turbulence
cannot be sustained when the cooling rate or  
turbulent viscosity is too large, resulting in disk fragmentation.     
%} 
We quantify the interpretation of this result as the
compromise of gravitational stability by: 
\begin{inparaenum}[1)] 
\item 
cooling, which reduces pressure support; and/or
\item 
viscosity, which reduces rotational support. 
\end{inparaenum}
We analyze the gravitational stability of accretion disks 
including these non-ideal effects explicitly.    
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal and
external heating, and cooling. We 
consider parameterized cooling functions in two and three dimensions,
as well as full radiative diffusion in 3D. 
We apply linear theory to 
global models of protoplanetary disks (PPDs) subject to gravito-turbulence modeled as an effective
viscosity and cooling via dust opacity. We find the sole effect of cooling is insufficient to
destabilize a realistic PPD. However, inclusion of viscosity renders
the disk beyond $\sim 60$AU dynamically unstable on radial 
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
%We suggest that realistic disk fragmentation can be
%physically attributed,  
%at least partially, to the removal of rotational stabilization against
%self-gravity by turbulence.   
We suggest turbulent stresses play an active role in realistic disk
fragmentation by removing rotational stabilization against 
self-gravity. 
%}
\end{abstract}
%we suggest turbulent stresses play an active, dynamical role in disk fragmentation by 
%removing rotational stabilization against self-gravity. 
%in the sense of small-scale transport

\section{Introduction}\label{intro} 
The gravitational instability (GI) of rotating disks    
plays a fundamental role in many astrophysical problems. In the context of 
gaseous protoplanetary disks (PPDs), GI has two applications. It can  
provide gravitational torques to transport angular momentum outwards
and thus enable mass accretion inwards \citep{armitage10,turner14}.   
GI may also lead to disk fragmentation, which has been invoked to 
explain giant planet formation at large radii \citep{helled14}. Any 
realistic study of these phenomena require direct   
numerical simulations. Nevertheless, physical insight can be  
obtained through analytical understanding of GI.  %physical insight

The standard metric for the (inverse) strength of disk self-gravity is the 
\emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre $Q$ is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). When $Q\leq 1$ the disk is formally unstable to local,
axisymmetric disturbances in a two-dimensional (2D, razor-thin),
inviscid and adiabatic (including isothermal) disk.

@@

The Toomre parameter is widely applied to assess GI in more general 
situtations (cf. the ideal disk models from which $Q$ is derived
from). This includes realistic disk structures 
that account for non-ideal physics not 
explicitly reflected by $Q$ \citep[e.g.][]{kimura12}.  It is, however,
not obvious that such additions leave the stability condition
unaltered.

Indeed, we show that when cooling and viscosity are  
included, as expected in realistic disks, GI cannot be assessed from a
single `Toomre-like' condition.      
%disk structure may be computed with sophisticated physics but
%stability condition still based on ideal disk models. 
%three main ingredients - gravity, pressure, rotation 
%a standard way to apply this condition is to compute the disk
%structure  based on realistic physics 
The goal of this work is to generalize the analytic treatment of disk 
GI to allow a self-consistent discussion of GI in realistic accretion 
disk models. 
%Specifically, we will examine the effect of cooling and  
%viscosity on GI in PPDs.

%Their relevance to PPDs are described below.

### 2016-01-21 paper/results3d.tex
reg udpate

%take k>0 wlog
%consider Gamma=1
%height s.t. rho(zmax) = 0.05 rho(0) avoid very low densities 
\subsection{Numerical procedure}
{\bf move to appendix?}
We use a pseudo-spectral method to solve the set of ordinary
differential equations, Eq. \ref{lin_mass}---\ref{lin_gravity}, on 
the domain $ z\in[0,\zmax]$ with a parity condition at the midplane.
We expand $\bm{U}=[\delta P,\delta\rho,\delta\Phi,
  \delta v_x, \delta 
  v_y]$ in even Chebyshev polynomials,
\begin{align}
  \bm{U}(z) = \sum_{j=1}^N \bm{a}_jT_{2(j-1)}(z/\zmax), 
\end{align}
and the vertical velocity in odd Chebyshev polynomials,
\begin{align}
  \delta v_z(z) = \sum_{j=1}^N b_jT_{2j-1}(z/\zmax),   
\end{align}
where $\bm{a}_j$ and $b_j$ are the spectral coefficients. 
The basis functions are chosen to satisfy a reflecting boundary
condition at the midplane, %GI has even symmetry across midplane 
\begin{align}
  \bm{U}^\prime(0) = \bm{0}, \quad \delta v_z(0) = 0.
\end{align}
%At the upper disk boundary we also apply reflection,
We explicitly apply a reflecting upper disk boundary, 
\begin{align}
  \delta v_z(\zmax) = \delta v_x^\prime(\zmax) = \delta
  v_y^\prime(\zmax) = 0,
\end{align}
and the potential satisfies
\begin{align}
  \delta \Phi^\prime (\zmax) + k \delta\Phi(\zmax) = 0, 
\end{align}
as derived by \cite{goldreich65a} and used in similar studies  
\citep{kim12,lin14c}. %reflection is standard in direct simulations

We discretize the equations, including upper disk boundary conditions,
over the $N$ positive abscissae of the extrema of $T_{2N-1}$. This
procedure converts the differential equations into a generalized
eigenvalue problem, for which we use the standard matrix package
LAPACK to solve.

>>>>

%take k>0 wlog
%consider Gamma=1
%height s.t. rho(zmax) = 0.05 rho(0) avoid very low densities
